 

### 2021-08-02 vertical-thermal-review.Rmd
Stepping into Warming section

**Vertical gradients affect tree metabolism and growth.**
Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
The proportion of GPP allocated to many of these increases with tree size, including ... foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].As a result of increasing C allocation to these other functions, mass-specific - and sometimes even absolute - woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021; @anderson-teixeiraCarbonCyclingMature2021].
However, NSC reserves increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

**Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in whole-tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021]. This is consistent with the observation that $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2). More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive] and lower capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020]. This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table 1, @mcdowellRelationshipTreeHeight2011].

The interannual climate sensitivity of annual woody growth can be determined through analysis of tree-ring increments.
While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions [@frittsTreeRingsClimate1976], only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [**Fig. `r fig_treerings`**, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021; @anderson-teixeiraJointEffectsClimate2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. In addition to lower drought resistance of growth, larger trees also frequently exhibit lower ability to bounce back from stress, ie. resilience [@mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Despite the potential for shorter trees in open forests to experience greater environmental stress [**REFS**], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].

@@

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees. In column  (a), large trees have a more negative response to high temperatures during the current or previous growing season. Shown are tree ring width (RW) responses to the most influential temperature variable at the site (Tmax or PET) of the largest versus smallest trees included in GLS model with a DBH x temperature interaction (Anderson-Teixeira et al. , in review) . In column (b), understory trees have more negative responses to high growing season temperatures across eight New England forests, (@rollinsonClimateSensitivityUnderstory2021) [ISSUE # 29](https://github.com/EcoClimLab/vertical-thermal-review/issues/29).](tree_rings/Figure_4_tree_rings.png)

@@

**Canopy trees account for the majority of forest ecosystem water and carbon cycling.**
While studies partitioning transpiration across forest vertical gradients are rare, both ET and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011]. Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration (Fig 5a). In terms of C cycling, it has been estimated that canopy strata contribute ???64% of net daytime $CO_{2}$ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007]. Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees ???10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests (Fig. 5b, Piponiot et al., accepted).

@@

**It is less clear how thermal sensitivity of water and C fluxes varies across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.** We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures-- at least under conditions of moderate to high VPD. A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an El Nino drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory. It remains far less clear how thermal sensitivity is likely to vary across forest strata under wet conditions, but both physiological mechanisms (Fig. 3) and tree ring evidence (Fig. 4) suggests that understory trees may respond more negatively to hot, wet conditions. Further research is required to better understand the thermal sensitivity of forest ecosystem function across strata.

>>>>

**Vertical gradients affect tree metabolism and growth.**
Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
The proportion of GPP allocated to many of these increases with tree size, including ... foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020]. As a result of increasing carbon allocation to these other functions, mass-specific - and sometimes even absolute - woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraCarbonCyclingMature2021].
However, NSC reserves increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

**Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in whole-tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021]. This is consistent with the observation that $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2). More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020]. This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table 1, @mcdowellRelationshipTreeHeight2011].

The interannual climate sensitivity of annual woody growth can be determined through analysis of tree-ring increments.
While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions [@frittsTreeRingsClimate1976], only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [**Fig. `r fig_treerings`**, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. In addition to lower drought resistance of growth, larger trees also frequently exhibit lower ability to bounce back from stress, ie. resilience [@mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].

@@

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees. In column (a), large trees have a more negative response to high temperatures during the current or previous growing season. Shown are tree ring width (RW) responses to the most influential temperature variable at the site ($T_{max}$ or PET) of the largest versus smallest trees included in GLS model with a DBH x temperature interaction (Anderson-Teixeira et al., in review). In column (b), understory trees have more negative responses to high growing season temperatures across eight New England forests, (@rollinsonClimateSensitivityUnderstory2021). 
[ISSUE # 29](https://github.com/EcoClimLab/vertical-thermal-review/issues/29).](tree_rings/Figure_4_tree_rings.png)

@@

**Canopy trees account for the majority of forest ecosystem water and carbon cycling.**
While studies partitioning transpiration across forest vertical gradients are rare, both ET and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011]. Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration (Fig 5a). In terms of C cycling, it has been estimated that canopy strata contribute $\geq$64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007]. Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$10cm DBH usually contributing $>$85% of ANPPwoody and Mwoody across 25 globally distributed forests (Fig. 5b, Piponiot et al., accepted).

@@

**It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.** We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures-- at least under conditions of moderate to high VPD. A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an El Nino drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory. It remains far less clear how thermal sensitivity is likely to vary across forest strata under wet conditions, but both physiological mechanisms (Fig. 3) and tree ring evidence (Fig. 4) suggests that understory trees may respond more negatively to hot, wet conditions. Further research is required to better understand the thermal sensitivity of forest ecosystem function across strata.
